Python as a machine learning language: experiences from the scikit-learn project

### 2013-06-12 paper.tex
New structure

In a previous paper \citep{pedregosa2011}, \sklearn was briefly presented and
benchmarked against several competitors. In this paper, we instead focus on an
in-depth discussion and analysis of the design choices used in
building the library. First, in section \ref{sec:api}, we present the
central application programming interface (API). Then, in section
\ref{sec:implementation}, we describe the implementation trade-offs.
In section \ref{sec:weaknesses}, we analyze the limitations of the
technologies we rely on; and finally, in section \ref{sec:comparison}, we
compare \emph{scikit-learn} to other major projects in terms of API.

\section{API}

\label{sec:api}

In this section, we discuss the general principles of the application
programming interface (API) of \sklearn. We first specify the
representation of data and then proceed by describing the interface shared by
all objects of the library. We then discuss the composition and reusability
advantages of that interface.

The API is designed to adhere to the following broad principles:
\begin{description}
  \item[Inspection.]
       Constructor parameters
       and parameter values determined by learning algorithms
       are stored as public attributes. All of them are documented in a consistent manner.
       % Gilles: Do we really need to bother the reader with such details?
       % (The Python language does not actually distinguish
       % between public and private attributes of objects,
       % but by convention, any identifier
       % prefixed with an underscore (\texttt{\_})
       % is considered ``a non-public part of the API''; \citealp{pythontut}.)
  \item[Non-proliferation of classes.]
       Learning algorithms are objects of custom classes,
       but datasets and hyperparameter names and values
       are represented as standard Python, NumPy or SciPy classes
       rather than new ones whenever feasible.
       % Gilles: I don't understand the phrase below
       % Custom classes adhere to Python conventions and protocols
       % for constructs such as iteration over the folds
       % in a $k$-fold cross validation.
       This should keep \sklearn easy to use
       and easy to combine with other libraries.
       % XXX do we need to elaborate more on this, or do we expect everyone
       % to be an experienced Java/adapter pattern hater?
  \item[Composition.]
       Many machine learning tasks are expressible
       as sequences or combinations of transformations to data.
       Some learning algorithms are also naturally viewed
       as meta-algorithms parameterized on other algorithms.
       To be able to implement and use these algorithms in a simple way,
       all learners (basic or composed) are usable with a consistent set of methods.
       % Gilles: let elaborate later on the syntax
       % and a simple syntax is implemented
       % for setting and retrieving parameters of sub-estimators.
       % XXX need to describe this!
  \item[Sensible defaults.]
       Whenever an operation requires a user-defined parameter,
       a default and appropriate value is defined by the library.
       The default value should cause the operation to be performed
       in a sensible way (giving a baseline solution for the task at hand).
\end{description}

@@

The public API is oriented towards batch processing
rather than processing single samples per API call.
While classification and regression algorithms can indeed make predictions
for single samples,
\sklearn objects are not optimized for this use case.
(The few online learning algorithms implemented
are intended to take minibatches.)
Batch processing makes optimal use of NumPy and SciPy
by preventing the overhead inherent to Python function calls
and per-element dynamic type checking.
Although this might seem to be an artifact of the Python language,
and therefore an implementation detail that leaks into the API,
we argue that APIs should be designed
so as to not tie a library to a suboptimal implementation strategy,
and batch processing can enable fast implementations
in lower-level languages as well
(where memory hierarchy effects
and the possibility of internal parallelization come into play).
% Gilles: I don't think it is worth elaborating too much (if at all) on online learning. They are exceptions in scikit-learn.

\subsection{Estimators, predictors and transformers}

All objects within \sklearn share a uniform common basic API
consisting of three complementary interfaces: an \textit{estimator} interface
for building and fitting objects, a \textit{predictor} interface for making
predictions and a \textit{transformer} interface for converting data.
As much as possible, our design choices have been guided so as to avoid
the proliferation of framework code. We try to adopt simple conventions and
to limit to a minimum the number of methods an object must implement.

\textbf{Estimators.} The \textit{estimator} interface is at the core of the
library. It defines instantiation mechanisms of objects and exposes a
\texttt{fit} method for learning a model from training data.  All supervised and
unsupervised learning algorithms (e.g., for classification, regression or
clustering) are offered as objects implementing that interface. Machine
learning tasks like feature extraction, feature selection or dimensionality
reduction are also provided as estimators.

@@

\textbf{Predictors.} The \textit{predictor} interface defines a \texttt{predict}
method taking as input some new data \texttt{X\_test} and producing as output
predictions for \texttt{X\_test}, based on the learned parameters of the
estimator (we call the input to \texttt{predict} ``\texttt{X\_test}'' in order
to emphasize that \texttt{predict} generalizes to new data). In the case of
supervised learning estimators, this method typically returns the predicted
labels or values computed by the model.  Continuing with the previous example,
predicted labels for \texttt{X\_test} can be obtained using the following
snippet:
\begin{verbatim}
    y_pred = clf.predict(X_test)
\end{verbatim}
Some unsupervised learning
estimators may however also implement that interface, as illustrated below:
\begin{verbatim}
    from sklearn.cluster import KMeans
    km = KMeans(n_clusters=10)
    km.fit(X_train)
    clust_pred = km.predict(X_test)
\end{verbatim}
In the above example, a \texttt{KMeans} estimator is first fit on the training
data \texttt{X\_train}. Then the \texttt{predict} method is called on the test
data \texttt{X\_test} in order to retrieve predicted cluster labels.
The \texttt{predict} method in \sklearn thus offers a unified way
to compute predictions with respect to unseen data.

@@

\textbf{Transformers.} Since it is common to modify or filter data before
feeding it to a learning algorithm, some estimators in the library implement a
\textit{transformer} interface which defines a \texttt{transform} method. It
takes as input some new data \texttt{X\_test} and yields as output a transformed
version of \texttt{X\_test}.
Preprocessing, feature selection and dimensionality
reduction algorithms are all provided as transformers within the library.  In
our example, if one wanted to preprocess the inputs \texttt{X\_train} before
fitting the logistic regression estimator, e.g., by standardizing the feature
values, one could thus simply write:
\begin{verbatim}
    from sklearn.preprocessing import StandardScaler
    scaler = StandardScaler()
    scaler.fit(X_train)
    X_train = scaler.transform(X_train)
\end{verbatim}
Of course, in practice, it is important to apply the same preprocessing to the
test data \texttt{X\_test}. This can easily be done as follows:
\begin{verbatim}
    X_test = scaler.transform(X_test)
\end{verbatim}
Transformers also include a variety of learning algorithms, such as
dimension reduction (PCA, manifold learning techniques), kernel
approximation techniques, and any other approach that can be used to
derive a modified representation of the data.

\subsection{Method chaining}

In \sklearn, the \texttt{fit} method always returns the estimator
it was called on, hence making method chaining possible.  Using this technique,
we can thus rewrite the \texttt{StandardScaler} example above in a single line:
\begin{verbatim}
    X_train = StandardScaler().fit(X_train).transform(X_train)
\end{verbatim}
In addition, \texttt{fit(X\_train).transform(X\_train)} can further
be rewritten as \texttt{fit\_transform(X\_train)}.  Combining the \texttt{fit}
and \texttt{transform} methods into a single method \texttt{fit\_transform}
allows to share computations, which often makes a huge difference in practice.
In the same spirit, \sklearn sometimes provides the
\texttt{fit\_predict} method for combining the \texttt{fit} and \texttt{predict}
methods. This is especially useful for retrieving cluster predictions with
respect to the training data \texttt{X\_train} in clustering algorithms.

\subsection{Estimator composition}

@@

\section{Weaknesses of the NumPy/SciPy environment}

\label{sec:weaknesses}

\textcolor{red}{TODO: move this into the Implementation section? These are mostly implementation considerations.}

@@

\section{Future work}

>>>>

In a previous paper \citep{pedregosa2011}, \sklearn was briefly presented and
benchmarked against several competitors. In this paper, we instead focus on an
in-depth discussion and analysis of the design choices used in building the
library. First, in section \ref{sec:core-api}, we present the central application
programming interface (API) and then describe, in section \ref{sec:advanced-api},
advanced API mechanisms built on this core API. Section
\ref{sec:implementation} describes the implementation trade-offs and analyzes
the limitations of the technologies we rely on while section
\ref{sec:comparison} then compares \emph{scikit-learn} to other major projects in
terms of API. Finally, conclusions and directions of future works are included
in section \ref{sec:conclusions}.

\section{Core API}

\label{sec:core-api}

In this section, we discuss the general principles of the application
programming interface (API) of \sklearn. We first specify the representation of
data and then proceed by describing the core interface of the library.

All objects within \sklearn share a uniform common basic API consisting of three
complementary interfaces: an \textit{estimator} interface for building and
fitting objects, a \textit{predictor} interface for making predictions and a
\textit{transformer} interface for converting data. As much as possible, our
design choices have been guided so as to avoid the proliferation of framework
code. We try to adopt simple conventions and to limit to a minimum the number of
methods an object must implement. The API is designed to adhere to the following
broad principles:

\begin{description}
  \item[Inspection.]
       Constructor parameters
       and parameter values determined by learning algorithms
       are stored as public attributes. All of them are documented in a consistent manner.
       % Gilles: Do we really need to bother the reader with such details?
       % (The Python language does not actually distinguish
       % between public and private attributes of objects,
       % but by convention, any identifier
       % prefixed with an underscore (\texttt{\_})
       % is considered ``a non-public part of the API''; \citealp{pythontut}.)
  \item[Non-proliferation of classes.]
       Learning algorithms are objects of custom classes,
       but datasets and hyperparameter names and values
       are represented as standard Python, NumPy or SciPy classes
       rather than new ones whenever feasible.
       % Gilles: I don't understand the phrase below
       % Custom classes adhere to Python conventions and protocols
       % for constructs such as iteration over the folds
       % in a $k$-fold cross validation.
       This should keep \sklearn easy to use
       and easy to combine with other libraries.
       % XXX do we need to elaborate more on this, or do we expect everyone
       % to be an experienced Java/adapter pattern hater?
  \item[Composition.]
       Many machine learning tasks are expressible
       as sequences or combinations of transformations to data.
       Some learning algorithms are also naturally viewed
       as meta-algorithms parameterized on other algorithms.
       To be able to implement and use these algorithms in a simple way,
       all learners (basic or composed) are usable with a consistent set of methods.
       % Gilles: let elaborate later on the syntax
       % and a simple syntax is implemented
       % for setting and retrieving parameters of sub-estimators.
       % XXX need to describe this!
  \item[Sensible defaults.]
       Whenever an operation requires a user-defined parameter,
       a default and appropriate value is defined by the library.
       The default value should cause the operation to be performed
       in a sensible way (giving a baseline solution for the task at hand).
\end{description}

@@

The public API is oriented towards batch processing rather than processing
single samples per API call. While classification and regression algorithms can
indeed make predictions for single samples, \sklearn objects are not optimized
for this use case. (The few online learning algorithms implemented are intended
to take minibatches.) Batch processing makes optimal use of NumPy and SciPy by
preventing the overhead inherent to Python function calls or due to per-element
dynamic type checking. Although this might seem to be an artifact of the Python
language, and therefore an implementation detail that leaks into the API, we
argue that APIs should be designed so as not to tie a library to a suboptimal
implementation strategy. As such, batch processing enable fast implementations
in lower-level languages (where memory hierarchy effects and the possibility of
internal parallelization come into play).

\subsection{Estimators}

The \textit{estimator} interface is at the core of the
library. It defines instantiation mechanisms of objects and exposes a
\texttt{fit} method for learning a model from training data.  All supervised and
unsupervised learning algorithms (e.g., for classification, regression or
clustering) are offered as objects implementing that interface. Machine
learning tasks like feature extraction, feature selection or dimensionality
reduction are also provided as estimators.

@@

\subsection{Predictors}

The \textit{predictor} interface defines a \texttt{predict}
method taking as input some new data \texttt{X\_test} and producing as output
predictions for \texttt{X\_test}, based on the learned parameters of the
estimator (we call the input to \texttt{predict} ``\texttt{X\_test}'' in order
to emphasize that \texttt{predict} generalizes to new data). In the case of
supervised learning estimators, this method typically returns the predicted
labels or values computed by the model.  Continuing with the previous example,
predicted labels for \texttt{X\_test} can be obtained using the following
snippet:
\begin{verbatim}
    y_pred = clf.predict(X_test)
\end{verbatim}
Some unsupervised learning
estimators may however also implement that interface, as illustrated below:
\begin{verbatim}
    from sklearn.cluster import KMeans
    km = KMeans(n_clusters=10)
    km.fit(X_train)
    clust_pred = km.predict(X_test)
\end{verbatim}
In the above example, a \texttt{KMeans} estimator is first fit on the training
data \texttt{X\_train}. Then the \texttt{predict} method is called on the test
data \texttt{X\_test} in order to retrieve predicted cluster labels.
The \texttt{predict} method in \sklearn thus offers a unified way
to compute predictions with respect to unseen data.

@@

\subsection{Transformers}

Since it is common to modify or filter data before feeding it to a learning
algorithm, some estimators in the library implement a \textit{transformer}
interface which defines a \texttt{transform} method. It takes as input some new
data \texttt{X\_test} and yields as output a transformed version of
\texttt{X\_test}. Preprocessing, feature selection and dimensionality reduction
algorithms are all provided as transformers within the library.  In our example,
if one wanted to preprocess the inputs \texttt{X\_train} before fitting the
logistic regression estimator, e.g., by standardizing the feature values, one
could thus simply write:
\begin{verbatim}
    from sklearn.preprocessing import StandardScaler
    scaler = StandardScaler()
    scaler.fit(X_train)
    X_train = scaler.transform(X_train)
\end{verbatim}
Of course, in practice, it is important to apply the same preprocessing to the
test data \texttt{X\_test}. This can easily be done as follows:
\begin{verbatim}
    X_test = scaler.transform(X_test)
\end{verbatim}
Transformers also include a variety of learning algorithms, such as
dimension reduction (PCA, manifold learning techniques), kernel
approximation techniques, and any other approach that can be used to
derive a modified representation of the data.

Finally, let us also note that, the \texttt{fit} method always returns the estimator
it was called on, hence making method chaining possible.  Using this technique,
we can thus rewrite the \texttt{StandardScaler} example above in a single line:
\begin{verbatim}
    X_train = StandardScaler().fit(X_train).transform(X_train)
\end{verbatim}
In addition, \texttt{fit(X\_train).transform(X\_train)} can further
be rewritten as \texttt{fit\_transform(X\_train)}.  Combining the \texttt{fit}
and \texttt{transform} methods into a single method \texttt{fit\_transform}
allows to share computations, which often makes a huge difference in practice.
In the same spirit, \sklearn sometimes provides the
\texttt{fit\_predict} method for combining the \texttt{fit} and \texttt{predict}
methods. This is especially useful for retrieving cluster predictions with
respect to the training data \texttt{X\_train} in clustering algorithms.

\section{Advanced API}

\label{sec:advanced-api}

Building on the core interface introduced in the previous section, we now
present in this section advanced API mechanisms for i) building meta-estimators,
ii) composing complex estimators and iii) selecting models.

\subsection{Meta-estimators}

\textcolor{red}{TODO}

\subsection{Pipelines and feature unions}

@@

%\section{Weaknesses of the NumPy/SciPy environment}

\textcolor{red}{TODO: The following corresponds to the old weaknesses section. It should be properly rewritten.}

@@

\section{Conclusions and future works}
\label{sec:conclusions}
